How to Avoid OOTA Without Really Trying

### 2024-09-03 paper/ootavacuous.tex
asplos: Remove Section 7 ("Real-World Constraints")
Summarize at point of references, as needed.  Move "Semantic Dependencies
and Tooling" to the "Issues and Refinements" section for further
summarization.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	See Section~\ref{sec:Hardware Architecture and Design}
	for more discussion about hardware speculation.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

\section{Real-World Constraints}
\label{sec:Real-World Constraints}

The C++ standard imposes many constraints, and these have been considered
in prior work.
In real-world C++ implementations, however, the standard's abstract machine
must be mapped onto real hardware, and this imposes additional
constraints stemming from hardware architecture and design.
This mapping and these constraints have been only partially accounted
for in the standard and other work.
Part of the reason for this is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was
released~\cite{JadeAlglave2011ppcmem,Maranget2012TutorialARMPower}.
But they do exist now.

The following sections discuss these constraints, starting with
hardware constraints and continuing with additional constraints imposed
by the standard.
An additional section discusses the OOTA-cycle implications for C++
tooling.

\subsection{Hardware Architecture and Design}
\label{sec:Hardware Architecture and Design}

\emph{Can we summarize this section, especially given that the ASPLOS
audience can be expected to have a greater understanding of these issues
than the C++ standards committee?}

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see why, keep in mind that on real systems observable effects
must eventually be committed, but no effect can be committed
while any of the computations that determine the effect remain
speculative.
Combining this with the obvious fact that a machine instruction cannot
commit until it is no longer speculative, we can draw two conclusions:
\begin{itemize}
\item	A load cannot commit until the store it reads from has
	committed.\footnote{
	An early ARMv8 memory model did allow loads that read from an
	earlier store in the same thread to commit before the store.
	This apparent exception merely reflects a difference in
	nomenclature; in the memory model a store was said to commit at
	the time when it made its new value available to other
	threads, whereas we say that a store commits at the time when
	the CPU has irrevocably decided that it will take place with a
	particular fully determined value and address.}
\item	A store cannot commit until all the loads on which it has a
	hardware dependency have committed.
\end{itemize}

As an example showing what can go wrong when these principles are
violated, consider Listing~\ref{lst:Speculated Store and Non-Speculated Load}.
Suppose that the storage for \co{x} is located immediately before that
for \co{y[]}, and
suppose further that hardware speculation incorrectly guesses the value
of \co{r1} will be $-1$, so that the speculated store of 42 on line~11 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then if line~13 reads from the speculative store and commits, it will load
the value 42 into \co{r2}---an observable effect since \co{r2}
is mentioned in the \co{exists} clause on line~16---although
by the rules of the C++ abstract machine the value must instead be zero.

\begin{listing}[tbp]
@@ DisplayLitmus litmus/speculative-store.litmus @@
\caption{Speculated Store and Non-Speculated Load}
\label{lst:Speculated Store and Non-Speculated Load}
\end{listing}

Applied to Listing~\ref{lst:OOTA Cycle} on
page~\pageref{lst:OOTA Cycle}, for another example,
the principles dictate that if the load on line~8 reads from the store
on line~18 and the load on line~16 reads from the store on line~10
(as they do in the execution described by the listing's \co{exists} clause),
then each of the stores must commit before its corresponding load.
A further application says that
neither line~10 nor line~12 can commit until after the load
in line~8 (and also the conditional on line~9) has committed.
Similarly, neither line~18 nor line~20 can commit until after line~16
(and~17) has committed.
Taken together, these constraints imply that the OOTA cycle in the
listing cannot be realized, because to do so would require that
none of the machine instructions corresponding to lines~8, 10, 16,
and~18 could commit before the others.
Furthermore, it's not possible to circumvent this reasoning by suggesting
that some of those lines could commit at the same time,
because instructions take time to execute even when executing
speculatively.

Yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent such
speculative cycles from committing.
As a special case of this requirement, if a load obtains
a value from a speculated store that has not yet committed,
and the store gets squelched,
then the load must be restarted before it commits even if the load and
the store are executed by different CPUs.

The requirement applies within a single
multithreaded core as well as between cores.
The fact that intracore communication is faster than intercore
communication does not magically give loads the ability to commit
before the stores they read from, no matter what cores they execute on.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-2.litmus @@
\caption{Causality Test Case 2}
\label{lst:Causality Test Case 2}
\end{listing}

Another feature we can expect of upcoming computer systems is the
addition of new hardware instructions.
For example, one could imagine a conditional-store instruction,
similar to existing conditional-move instructions but carrying out
a store to memory rather than a move to a register.
The condition would control whether or not the store takes place.
A compiler doing single-thread analysis could use a conditional-store
instruction to generate code corresponding to lines~10 and~11 of
Listing~\ref{lst:Causality Test Case 2}
on page~\pageref{lst:Causality Test Case 2}
without use of a conditional branch, potentially
enabling more hardware optimizations than would otherwise be permitted.
Nevertheless, the conditional store could not be committed until
the instructions it depends on (lines~8 and~9) have committed.
Any new instruction will be subject to this constraint,
just as the instructions in current systems are.

A key conclusion to draw from this discussion is that
regardless of how advanced a computer system may be,
it cannot commit a store having a hardware dependency on some
set of loads until after (in global time) those loads have committed.

@@

Despite these outstanding issues, our definition of semantic dependency
allows us to bring to bear the real-world constraints outlined in the
next section.

>>>>

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.
The need to prohibit simple OOTA is one reason why compiler-based value
speculation optimizations require checks on such speculation, and further
requires that these checks must be based on the values actually loaded.

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

@@
